### 2021-06-11 papers/thesis/full/ch4-spherical-caps/spherical-caps.tex
Further edits to address comments. Some questions remain.

A continuation to the procedure for defining OPs on quadratic curves (see \cite{olver2021orthogonal}) has also been presented \cite{olver2020orthogonal}, where OPs are defined on quadratic surfaces of revolution in higher dimensions (specifically, surfaces in $\R^{d+1} for d \ge 2$). The authors describe their technique as modelled on the structure of the unit sphere, by approaching the construction as a generalisation of spherical harmonics.

@@

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\bseqn{
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:sc:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\bseqn{
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,y,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$ and thus its spherical gradient would be zero on the boundary $z = \alpha$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in \bsreffig{fig:sc:sparsity}.

>>>>

A continuation to the procedure for defining OPs on quadratic curves (see \cite{olver2021orthogonal}) has also been presented \cite{olver2020orthogonal}, where OPs are defined on quadratic surfaces of revolution in higher dimensions (specifically, surfaces in $\R^{d+1}$ for $d \ge 2$). The authors describe their technique as modelled on the structure of the unit sphere, by approaching the construction as a generalisation of spherical harmonics.

@@

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, \bstodo{All functions not dependent on r (ie any we expand in this basis) satisfy the derivative BC automatically. I am fairly sure this is still the BC for the sphere though..?}
\bseqn{
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:sc:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\bseqn{
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,y,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$ and thus its spherical gradient would be zero on the boundary $z = \alpha$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in \bsreffig{fig:sc:sparsity}.

### 2021-06-11 papers/thesis/full/conclusion/summary.tex
Further edits to address comments. Some questions remain.

In this thesis, we have developed sparse spectral methods for solving partial differential equations (PDEs) on disk-slices and trapeziums in 2D, and spherical caps as a surface 3D. The work can also be used as a template for developing similar methods on other such similar domains, in particular other subdomains of the sphere.

@@

For the disk-slice in 2D, we defined the OPs that allow similar sparse and banded-block-banded operators required for solving PDEs in the domain, deriving their structure and providing a method to efficiently calculate numerically their entries. The reason they need to be calculated numerically is due to the non-classical univariate OPs that are involved in the 2D OP definitions. Finally, we moved on to use the same arguments for the spherical cap, a 3D surface that is a subdomain of the unit sphere. We once again defined the 3D OPs as orthogonal polynomials in in $x,y,z$, and showed how the differential operator matrices continue to elicit similar banded-block-banded structure.

@@

Ideally, we would like to pair this with a spectral element method for the sphere, where the elements would be the spherical subdomains of spherical bands and spherical caps. The goal here would be to test this method against the full sphere spectral method the is in place in the ECMWF model, with the hypothesis being that by reducing the size of the transforms, we can improve the overall efficiency while still maintaining the accuracy that we achieve from a spectral approach.

>>>>

In this thesis, we have developed sparse spectral methods for solving partial differential equations (PDEs) on disk-slices and trapeziums in 2D, and spherical caps as a surface in 3D. The work can also be used as a template for developing similar methods on other such similar domains, in particular other subdomains of the sphere.

@@

For the disk-slice in 2D, we defined the OPs that allow similar sparse and banded-block-banded operators required for solving PDEs in the domain, deriving their structure and providing a method to efficiently calculate numerically their entries. The reason they need to be calculated numerically is due to the non-classical univariate OPs that are involved in the 2D OP definitions. Finally, we moved on to use the same arguments for the spherical cap, a 3D surface that is a subdomain of the unit sphere. We once again defined the 3D OPs as orthogonal polynomials in $x,y,z$, and showed how the differential operator matrices continue to elicit similar banded-block-banded structure.

@@

A further avenue to take the framework for scalar functions is to construct a vector polynomial basis for functions whose values lie on the tangent bundle of the sphere. In \bsrefappendix{appendix:tangent}, we provide a summary for how one needs to approach this. The motivation for such is so as to be able to derive similarly sparsely structured operators for the spherical gradient and divergence. The main observation however is that, while the vector spherical harmonics are already established as being such an orthogonal and complete basis for the whole sphere, a similar construction for the spherical cap would not yield the same result.

Ideally, we would like to pair the framework presented in \bsrefchapter{CHAPTER:sphericalcaps} with a spectral element method for the sphere, where the elements would be the spherical subdomains of spherical bands and spherical caps. The goal here would be to test this method against the full sphere spectral method the is in place in the ECMWF model, with the hypothesis being that by reducing the size of the transforms, we can improve the overall efficiency while still maintaining the accuracy that we achieve from a spectral approach.

Discretizations of partial differential operators are sparse on other suitable sub-domains of the sphere, as evidenced by the construction presented for the spherical cap. Hence, using the techniques learnt here, one could also look to constructing similar methods for solving PDEs on spherical triangles, with a view to using these sub-domains as elements for a spherical spectral element method. Moreover, one could use this framework to also develop sparse spectral methods on more general domains defined by an algebraic surface -- in particular those defined by non-uniform rational B-splines (NURBS) \bstodo{Not sure what to say exactly or how in depth to go here}. However, it is not obvious how one would derive or compute the OPs for such domains.
